What's in a Name? Linear Temporal Logic Literally Represents Time Lines

### 2023-06-26 paper/main.tex
minor changes
small changes

\section{Introduction}
Requirement specification is a central step in the development of safety-critical systems. As a first step, requirements are typically written in natural language. For example, here is a real-world requirement specification from an air traffic control system ~\cite{ZR14}
\begin{center}
    ``If a TSAFE command is sent to an aircraft, controller/AutoResolver should then hand off the control of this aircraft''
\end{center}
Such natural language requirements are often ambiguous and not amenable to formal analysis, so requirements must often be specified using rigorous formalized semantics for the purpose of verification and validation. The above requirement, when translated to Linear Temporal Logic (detailed in \cref{sec:ltl}), may be expressed as follows:
\begin{align*}
    \always & (\texttt{tsafe.TSAFE\_command1} \land \\ & \texttt{controller.CTR\_control\_1} \\
    \limplies & \nextt (\neg \texttt{controller.CTR\_control\_1}))
\end{align*}
Several tools have been developed to reason about formal specifications written with such \ltl\ formulas, like R2U2 and FRET \cite{GPMS20}. However, these formalized semantics are unintuitive and error-prone for human practitioners, and the tools are much less useful when system designers can't validate input specifications.
Thus, a significant hurdle remains: how can we convincingly demonstrate to the humans in the loop, from system designers to certifiers, that the analyzed formulas truly represent the desired system requirements? % quote MLTL paper?

@@

We must begin by setting up the prerequisite definitions involved in our work.
First, we provide the definition and semantics for \ltl. In the following subsections, we define (state-based) \Buchi automata and $\omega$-regular expressions, which are generalizations of the well-known deterministic finite automata (DFAs) and regular expressions, for the case of infinite words. Lastly, we describe the graphic visualizations we will use to depict timelines.

@@

\begin{remark}
    The tool we provide in this paper uses a set of concrete syntax to represent those logical connectives, in order to avoid the difficulty of typesetting connectives such as $\always$ and $\eventually$. The concrete syntax is defined in \ref{sec:concrete-syntax}.
\end{remark}

@@

\subsection{LTL to BA} \label{ltl2aut}
The automata-theoretic approach \cite{ORBi-a8d60ab6-1101-4434-9511-c01ea4e5a15b} of studying \ltl\ (i.e. reducing \ltl\ formulas to their corresponding \Buchi automata) has been well-studied. Many algorithms and optimizations have been studied such as \cite{DGV99, F03}. Our tool \tool uses SPOT~\cite{Dur22}\footnote{For our tool, we use SPOT version 2.11.} for this step, which provides an implementation of the translation from \ltl\ formulas to \Buchi automata.

@@

\begin{algorithm}[h!]
    \caption{nfa2regex}
    \begin{algorithmic}
        \Require ($G, s, f$), an NFA with initial state $s$ and final state $f$
        \Ensure The regular expression corresponding to all paths from $s$ to $f$
        % WHile there exists any edge except r_sf
        \While{ there exists an interior vertex $v$ }
            \State reduce\_nfa($G, v$)
            \State combine multiedges, i.e. $r1 : u \to w, r2 : u \to w$  to $r1 | r2 : u \to w$
        \EndWhile
        \State \Return $(r_\mathrm{ss}| r_\mathrm{sf} r_\mathrm{ff}^* r_\mathrm{fs})^* r_\mathrm{sf} r_\mathrm{ff}$
        % if called by first visit, this is r_\mathrm{ss}* r_\mathrm{sf}
        % if called by ba2wregex, s=f, this is r_\mathrm{ff}
        % consider modularizing fns differently
        % \State \Return regexp labelling the edge $s\to f$
    \end{algorithmic}
\end{algorithm}

@@

There is no agreed upon canonical form for regular expression, hence we do not hope to find the simplest regular expression for visualization. Regular expression simplification comes down to regular expression equivalence checking, which is computationally hard \cite{MNS04}. For purpose of simplification, one could perform a search over equivalent regular expressions and decide which one is simpler to use. However, this strategy is expensive in terms of the running time, and for the purpose of visualization, we did not use this strategy in our tool. We consider this approach as future work.

@@

        \resizebox{.8\textwidth}{!}{

@@

We gather a test suite of 91 formulas used in real-world requirement specification. We take $6$ specifications from NASA's Automated Airspace Concept (AAC)\cite{GCMTR16}. We also collect formulas from the Acacia suite of examples \cite{RV10}. We reap $85$ formulas from the suite of $23$ examples, by generating separate visualizations for each line of specification. We choose to visualize each formula specifically, rather than complete specifications, as complete specifications may describe large complex systems, and we only intend the use case of validating individual specification formulas. % We choose to visualize each formula specifically, rather than complete specifications, as

@@

\begin{figure}[h!]
    \centering
    \subfloat{{\includegraphics[width=6.5cm]{img/tllens-graph.png} }}%
    \quad % [\centering label 1]
    \subfloat{{\includegraphics[width=6.5cm]{img/star-height-graph.png}}}
    \caption{Visualization complexity metrics for formulas collected from Acacia \cite{RV10} and NASA's  AAC \cite{GCMTR16}}
    \label{fig:real-world}
\end{figure}

\paragraph{Scalability}
We measure the scalability of our tool by running it against a set of scalable \ltl\ formulas. We define a scalable \ltl\ formula as one that describes an $n$-bit number using a binary counter \cite{RV10}. As $n$ gets larger, the \ltl\ formula used to encode the $n$-bit counter scales in size, but it remains practical. The result is shown in \cref{fig:scalability}. In this analysis, the timeline length grows approximately exponentially with the number of bits in the counter, which matches the result in Rozier and Vardi \cite{RV10}.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[scale=0.6]{img/binary.png}
    \caption{Scalability measured with binary counter examples from Rozier and Vardi \cite{RV10}}
    \label{fig:scalability}
\end{figure}

% KYR: Here we characterize the size of the timeline as a function of the number of bits of a binary counter for each of the four binary counter \ltl\ formulas. The point is that these four are the only (unboundedly) scalable formulas that describe something real (i.e., they describe a binary counter for any number of bits) that have ever been published in the literature. These would throw off our metrics about the set of real \ltl\ formulas if we included them in there. Thus, we analyze them separately, and can define the timeline size as a function of the number of bits for each of the four formulas. We can give an example or two here, such as two pictures of timelines for a 2-bit and a 3-bit counter from two different formulas, one with U's and one with X's; one with carry and one without (such as a 2-bit w/carry X formula and a 3-bit no carry U formula). The scalable counter formulas are from \cite{RV10}.

@@

\section{Discussion}
Based on our results, we propose the following questions for discussions and future works:
\begin{itemize}
    \item Can there be more simplifications methods as described in Section \ref{regex-simplify}? The answer is definitely positive. On one hand, more rule-based simplifications can be discovered by finding more patterns in the generated $\omega$-regular expressions. On the other hand, other simplification are possible: for example, if the generated $\omega$-regular expression is $r_1 + r_2$ where $r_1, r_2$ are not syntactically equivalent (or equivalent up to algebraic rules) $\omega$-regular expressions, but they accepts the same set of infinite-length strings, then we can simplify this down to just $r_1$. This reduce the problem to equivalence test on $\omega$-regular expressions, which is decidable. In the similar vein, many more simplifications can be used, but the question is how efficient are they? In the end, one have to balance trade-off between the efficiency of the tool and the simplicity of the result.
    %\item Our tool is effective at translating a wide range of LTL formulas. Even with LTL formulae with complex nesting structure are represented by relatively readable timeline visualizations.
    \item Can we define timeline more formally so that one can do this process reversely (i.e. a tool that converts timeline representation to its corresponding \ltl\ formula)? That would be a very useful tool in practice. However, a direct reverse of the algorithm we presented in this paper may not be viable.
\end{itemize}

@@

\item {\bf What are intuitive representations of star-formulas and do they change depending on the formula?} Common regular expression visualizers use two ways to represent stars in regular expressions. We can represent a stared formula inside a box \cite{B22} with related caption to indicate the repetition (similar to the design decision we made in \tool); or with back edges \cite{A20}. Does using the box cause confusion with respect to the LTL $\Box$ operator? What is the most intuitive way to remind users of the corner case where the starred formula occurs zero times?

@@

\section{Conclusion}
The Achilles heel of formal verification is specification; formal methods are only as effective at verification as their specifications are at describing the essential properties to verify. Yet, specification remains the biggest bottleneck to the use of formal methods \cite{Roz16}. \ltl\ is one of the most popular specification logics for industrial-scale critical systems; in the space domain alone, it is currently encapsulating specifications for the development of the NASA Lunar Gateway \cite{DBR21}, the Air Force Research Laboratory/Collins Aerospace Spacecraft Collision Avoidance system \cite{HDWF21}, Space Systems Finland's Attitude and Orbit Control Systems (AOCS) \cite{ILLTV13}, and NASA/JPL's Europa Lander Mission Concept \cite{CDRWRWL22} just to name a few.
 %
Yet the humans that need to use formal verification tools and deeply understand their results struggle to validate that \ltl\ formulas capture the specifications they are meant to capture. A major contributor to \ltl's popularity is the propensity of humans to think of requirements in terms of timelines. This inspired the creation of \ltl\ in the first place, as a logic that ``intuitively'' represents timelines. Our work serves to reinforce that connection, enabling visualization of most realistic \ltl\ formulas as timelines. By releasing \tool, we contribute to better validation capabilities for \ltl\ specifications and aid the effort to make formal verification more accessible and wide-spread.
%We hope that this will assist system designers in formulating error-free specifications and thus improve the effectiveness of verification tools for \ltl.

@@

\appendix
\section{Concrete Syntax of \tool} \label{sec:concrete-syntax}
Recall from \ref{ltl-defn} that we define \ltl\ formula as
\begin{align*}
    \Phi ::= & p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi \\ & \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi
\end{align*}
where $p \in \AP$. In our tool \tool, we use the following concrete syntax to represent \ltl\ formulae.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \Phi & ::=  & p                                     & \texttt{p}                                     & \text{atomic proposition} \\
         & \mid & \neg \Phi                             & \texttt{!} \Phi                                & \text{negation} \\
         & \mid & \Phi \land \Phi                       & \Phi \texttt{ \& } \Phi                        & \text{conjunction} \\
         & \mid & \Phi \lor \Phi                        & \Phi \texttt{ | } \Phi                         & \text{disjunction} \\
         & \mid & \Phi \limplies \Phi                   & \Phi \texttt{ -> } \Phi                        & \text{implication} \\
         & \mid & \always \Phi                          & \texttt{G} \Phi                                & \text{globally} \\
         & \mid & \eventually \Phi                      & \texttt{F} \Phi                                & \text{finally} \\
         & \mid & \nextt \Phi                           & \texttt{X} \Phi                                & \text{next} \\
         & \mid & \Phi \stronguntil \Phi                & \Phi \texttt{ U } \Phi                         & \text{until} \\
         & \mid & \Phi \weakrelease \Phi                & \Phi \texttt{ R } \Phi                         & \text{release} \\
    \end{array}
\]
%\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{RelatedWork}

@@

\noindent
\begin{table}[h]
\begin{tabular}{p{3.6in}|c|c}
  \hline
  Purpose & \# formulas & Source \\
  \hline
  Design space for NASA's AAC (Automated Airspace Concept) & 20,250 & \cite{GCMTR16} \\
  \hline
  Safety specifications for Boeing's AIR6110 wheel brake system & & \cite{BCFJKPRT15} \\
  \hline
  Requirements for model checking NASA's AAC high-level architecture & 6 & \cite{ZR14} \\
  \hline
  Real formulas collected from all publicly-available, pre-2011 case studies & 63 & \cite{RV11} \\
  \hline
\end{tabular}
\end{table}

>>>>

\section{Introduction}
Requirement specification is a central step in the development of safety-critical systems. As a first step, requirements are typically written in natural language. For example, here is a real-world requirement specification from an air traffic control system~\cite{ZR14}
\begin{center}
    ``If a TSAFE command is sent to an aircraft, controller/AutoResolver should then hand off the control of this aircraft''
\end{center}
Such natural language requirements are often ambiguous and not amenable to formal analysis, so requirements must often be specified using rigorous formalized semantics for the purpose of verification and validation. The above requirement, when translated to Linear Temporal Logic (detailed in \cref{sec:ltl}), may be expressed as follows:
\begin{align*}
    \always & (\texttt{tsafe.TSAFE\_command1} \land \\ & \texttt{controller.CTR\_control\_1} \\
    \limplies & \nextt (\neg \texttt{controller.CTR\_control\_1}))
\end{align*}
Several tools have been developed to reason about formal specifications written with such \ltl\ formulas, like R2U2 and FRET~\cite{GPMS20}. However, these formalized semantics are unintuitive and error-prone for human practitioners, and the tools are much less useful when system designers can't validate input specifications.
Thus, a significant hurdle remains: how can we convincingly demonstrate to the humans in the loop, from system designers to certifiers, that the analyzed formulas truly represent the desired system requirements? % quote MLTL paper?

@@

We begin by setting up the prerequisite definitions involved in our work.
First, we provide the definition and semantics for \ltl. In the following subsections, we define (state-based) \Buchi automata and $\omega$-regular expressions, which are generalizations of the well-known deterministic finite automata (DFAs) and regular expressions, for the case of infinite words. Lastly, we describe the graphic visualizations we will use to depict timelines.

@@

\begin{remark}
    The tool we provide in this paper uses a set of concrete syntax to represent those logical connectives, in order to avoid the difficulty of typesetting connectives such as $\always$ and $\eventually$. The concrete syntax is defined in~\ref{sec:concrete-syntax}.
\end{remark}

@@

\subsection{LTL to BA} \label{ltl2aut}
The automata-theoretic approach~\cite{ORBi-a8d60ab6-1101-4434-9511-c01ea4e5a15b} of studying \ltl\ (i.e. reducing \ltl\ formulas to their corresponding \Buchi automata) has been well-studied. Many algorithms and optimizations have been studied such as~\cite{DGV99, F03}. Our tool \tool uses SPOT~\cite{Dur22}\footnote{For our tool, we use SPOT version 2.11.} for this step, which provides an implementation of the translation from \ltl\ formulas to \Buchi automata.

@@

\begin{algorithm}[h!]
    \caption{nfa2regex}
    \begin{algorithmic}
        \Require ($G, s, f$), an NFA with initial state $s$ and final state $f$
        \Ensure The regular expression corresponding to all paths from $s$ to $f$
        % WHile there exists any edge except r_sf
        \While{ there exists an interior vertex $v$ }
            \State reduce\_nfa($G, v$)
            \State combine multi-edges, i.e. $r1 : u \to w, r2 : u \to w$  to $r1 | r2 : u \to w$
        \EndWhile
        \State \Return $(r_\mathrm{ss}| r_\mathrm{sf} r_\mathrm{ff}^* r_\mathrm{fs})^* r_\mathrm{sf} r_\mathrm{ff}$
        % if called by first visit, this is r_\mathrm{ss}* r_\mathrm{sf}
        % if called by ba2wregex, s=f, this is r_\mathrm{ff}
        % consider modularizing fns differently
        % \State \Return regexp labelling the edge $s\to f$
    \end{algorithmic}
\end{algorithm}

@@

There is no agreed upon canonical form for regular expression, hence we do not hope to find the simplest regular expression for visualization. Regular expression simplification comes down to regular expression equivalence checking, which is computationally hard~\cite{MNS04}. For purpose of simplification, one could perform a search over equivalent regular expressions and decide which one is simpler to use. However, this strategy is expensive in terms of the running time, and for the purpose of visualization, we did not use this strategy in our tool. We consider this approach as future work.

@@

        \resizebox{.55\textwidth}{!}{

@@

We gather a test suite of 91 formulas used in real-world requirement specification. We take $6$ specifications from NASA's Automated Airspace Concept (AAC)~\cite{GCMTR16}. We also collect formulas from the Acacia suite of examples~\cite{RV10}. We reap $85$ formulas from the suite of $23$ examples, by generating separate visualizations for each line of specification. We choose to visualize each formula specifically, rather than complete specifications, as complete specifications may describe large complex systems, and we only intend the use case of validating individual specification formulas. % We choose to visualize each formula specifically, rather than complete specifications, as

@@

\begin{figure}[h!]
    \centering
    \subfloat{{\includegraphics[width=6.5cm]{img/tllens-graph.png} }}%
    \quad % [\centering label 1]
    \subfloat{{\includegraphics[width=6.5cm]{img/star-height-graph.png}}}
    \caption{Visualization complexity metrics for formulas collected from Acacia~\cite{RV10} and NASA's  AAC~\cite{GCMTR16}}
    \label{fig:real-world}
\end{figure}

\paragraph{Scalability}
We measure the scalability of our tool by running it against a set of scalable \ltl\ formulas. We define a scalable \ltl\ formula as one that describes an $n$-bit number using a binary counter~\cite{RV10}. As $n$ gets larger, the \ltl\ formula used to encode the $n$-bit counter scales in size, but it remains practical. The result is shown in \cref{fig:scalability}. In this analysis, the timeline length grows approximately exponentially with the number of bits in the counter, which matches the result in Rozier and Vardi~\cite{RV10}.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=6.5cm]{img/binary.png}
    \caption{Scalability measured with binary counter examples from Rozier and Vardi~\cite{RV10}}
    \label{fig:scalability}
\end{figure}

% KYR: Here we characterize the size of the timeline as a function of the number of bits of a binary counter for each of the four binary counter \ltl\ formulas. The point is that these four are the only (unboundedly) scalable formulas that describe something real (i.e., they describe a binary counter for any number of bits) that have ever been published in the literature. These would throw off our metrics about the set of real \ltl\ formulas if we included them in there. Thus, we analyze them separately, and can define the timeline size as a function of the number of bits for each of the four formulas. We can give an example or two here, such as two pictures of timelines for a 2-bit and a 3-bit counter from two different formulas, one with U's and one with X's; one with carry and one without (such as a 2-bit w/carry X formula and a 3-bit no carry U formula). The scalable counter formulas are from~\cite{RV10}.

@@

\section{Discussion}
Based on our results, we propose the following questions for discussions and future works:
\begin{itemize}
    \item Can there be more simplifications methods as described in Section~\ref{regex-simplify}? The answer is definitely positive. On one hand, more rule-based simplifications can be discovered by finding more patterns in the generated $\omega$-regular expressions. On the other hand, other simplification are possible: for example, if the generated $\omega$-regular expression is $r_1 + r_2$ where $r_1, r_2$ are not syntactically equivalent (or equivalent up to algebraic rules) $\omega$-regular expressions, but they accepts the same set of infinite-length strings, then we can simplify this down to just $r_1$. This reduce the problem to equivalence test on $\omega$-regular expressions, which is decidable. In the similar vein, many more simplifications can be used, but the question is how efficient are they? In the end, one have to balance trade-off between the efficiency of the tool and the simplicity of the result.
    %\item Our tool is effective at translating a wide range of LTL formulas. Even with LTL formulae with complex nesting structure are represented by relatively readable timeline visualizations.
    \item Can we define timeline more formally so that one can do this process reversely (i.e. a tool that converts timeline representation to its corresponding \ltl\ formula)? That would be a very useful tool in practice. However, a direct reverse of the algorithm we presented in this paper may not be viable.
\end{itemize}

@@

\item {\bf What are intuitive representations of star-formulas and do they change depending on the formula?} Common regular expression visualizers use two ways to represent stars in regular expressions. We can represent a stared formula inside a box~\cite{B22} with related caption to indicate the repetition (similar to the design decision we made in \tool); or with back edges~\cite{A20}. Does using the box cause confusion with respect to the LTL $\Box$ operator? What is the most intuitive way to remind users of the corner case where the starred formula occurs zero times?

@@

\section{Conclusion}
The Achilles heel of formal verification is specification; formal methods are only as effective at verification as their specifications are at describing the essential properties to verify. Yet, specification remains the biggest bottleneck to the use of formal methods~\cite{Roz16}. \ltl\ is one of the most popular specification logics for industrial-scale critical systems; in the space domain alone, it is currently encapsulating specifications for the development of the NASA Lunar Gateway~\cite{DBR21}, the Air Force Research Laboratory/Collins Aerospace Spacecraft Collision Avoidance system~\cite{HDWF21}, Space Systems Finland's Attitude and Orbit Control Systems (AOCS)~\cite{ILLTV13}, and NASA/JPL's Europa Lander Mission Concept~\cite{CDRWRWL22} just to name a few.
 %
Yet the humans that need to use formal verification tools and deeply understand their results struggle to validate that \ltl\ formulas capture the specifications they are meant to capture. A major contributor to \ltl's popularity is the propensity of humans to think of requirements in terms of timelines. This inspired the creation of \ltl\ in the first place, as a logic that ``intuitively'' represents timelines. Our work serves to reinforce that connection, enabling visualization of most realistic \ltl\ formulas as timelines. By releasing \tool, we contribute to better validation capabilities for \ltl\ specifications and aid the effort to make formal verification more accessible and wide-spread.
%We hope that this will assist system designers in formulating error-free specifications and thus improve the effectiveness of verification tools for \ltl.

@@

\appendix
\section{Concrete Syntax of \tool} \label{sec:concrete-syntax}
Recall from~\ref{ltl-defn} that we define \ltl\ formula as
\begin{align*}
    \Phi ::= & p \mid \neg \Phi \mid \Phi \land \Phi \mid \Phi \lor \Phi \mid \Phi \limplies \Phi \mid \always \Phi \mid \eventually \Phi \mid \nextt \Phi \\ & \mid \Phi \stronguntil \Phi \mid \Phi \weakrelease \Phi
\end{align*}
where $p \in \AP$. In our tool \tool, we use the following concrete syntax to represent \ltl\ formulae.
\[
    \begin{array}{c r l l l@{\qquad} l}
         &      & \textit{abstract syntax}              & \textit{concrete syntax}                       & \textit{description} \\
    \Phi & ::=  & p                                     & \texttt{p}                                     & \text{atomic prop.} \\
         & \mid & \neg \Phi                             & \texttt{!} \Phi                                & \text{negation} \\
         & \mid & \Phi \land \Phi                       & \Phi \texttt{ \& } \Phi                        & \text{conjunction} \\
         & \mid & \Phi \lor \Phi                        & \Phi \texttt{ | } \Phi                         & \text{disjunction} \\
         & \mid & \Phi \limplies \Phi                   & \Phi \texttt{ -> } \Phi                        & \text{implication} \\
         & \mid & \always \Phi                          & \texttt{G} \Phi                                & \text{globally} \\
         & \mid & \eventually \Phi                      & \texttt{F} \Phi                                & \text{finally} \\
         & \mid & \nextt \Phi                           & \texttt{X} \Phi                                & \text{next} \\
         & \mid & \Phi \stronguntil \Phi                & \Phi \texttt{ U } \Phi                         & \text{until} \\
         & \mid & \Phi \weakrelease \Phi                & \Phi \texttt{ R } \Phi                         & \text{release} \\
    \end{array}
\]
%\nocite{*}
\bibliographystyle{IEEEtran}
\bibliography{RelatedWork}

@@

\noindent
\begin{table}[h]
\begin{tabular}{p{3.6in}|c|c}
  \hline
  Purpose & \# formulas & Source \\
  \hline
  Design space for NASA's AAC (Automated Airspace Concept) & 20,250 &~\cite{GCMTR16} \\
  \hline
  Safety specifications for Boeing's AIR6110 wheel brake system & &~\cite{BCFJKPRT15} \\
  \hline
  Requirements for model checking NASA's AAC high-level architecture & 6 &~\cite{ZR14} \\
  \hline
  Real formulas collected from all publicly-available, pre-2011 case studies & 63 &~\cite{RV11} \\
  \hline
\end{tabular}
\end{table}
